Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2019-03-26 WholePlotTrajectories.tex
relecture

While both taxonomic and functional composition remained stable in
undisturbed communities (Fig. \ref{fig:NMDSplans}), they followed marked
and consistent trajectories over post-\break disturbance time. In
disturbed communities, these compositional changes corresponded to
shifts towards species with more acquisitive functional strategies, from
communities with high mean WSG to high mean SLA and chlorophyll content
(see appendix I). For functional composition, this translated into
cyclic compositional changes with an incomplete recovery of the initial
composition (Fig. \ref{fig:NMDSplans}). The maximum dissimilarity with
the initial state was positively correlated with the disturbance
intensity for both taxonomic and functional composition
(\(\rho_{Spearman}^{Taxonomic}=0.87\) and
\(\rho_{Spearman}^{Functional}=0.90\) respectively). The maximum
dissimilarity with the initial was reached for taxonomic composition
between 15 to 25 years, for most of the plots, and around 22 years for
functional composition.

@@

Community CWM average value of Maximum height at adult stage
(\emph{Hmax}), leaf toughness and wood specific gravity (\emph{WSG})
first decreased and then slightly increased but remained significantly
lower than their initial value (Fig. \ref{fig:CWM}). On the other side,
bark thickness and specific leaf area (\emph{SLA}) increased and while
bark thickness remained substantially high after 30 years, \emph{SLA}
had almost recovered to its initial value. For all traits, the maximum
difference to initial value was correlated to the disturbance intensity.
Positive correlations were observed for Leaf thickness, chlorophyll
content, SLA and bark thickness
(\(\rho_{Spearman}^{Leaf thickness}=0.76\),
\(\rho_{Spearman}^{Chlorophyll content}=0.60\),
\(\rho_{Spearman}^{SLA}=0.93\),
\(\rho_{Spearman}^{Bark thickness}=0.71\)). Negative correlation was
observed for Leaf toughness, WSG and Hmax
(\(\rho_{Spearman}^{Leaf toughness}=-0.53\),
\(\rho_{Spearman}^{WSG}=-0.75\), \(\rho_{Spearman}^{Hmax}=-0.40\)) The
proportions of the three lightest seed mass classes increased in all
disturbed plots. After 30 years the proportion of lightest seed mass
class decreased while it stabilized for the two other (Supp. Mat. - Fig.
S2).

@@

For undisturbed plots, taxonomic richness and Simpson diversity remained
stable over the 30 years of monitoring. In disturbed communities, after
low disturbance intensity the taxonomic richness increased, reaching a
maximum gain of 14 botanical genera (plot 3 from treatment 2). After
intense disturbance the taxonomic richness followed a more complex
trajectory, decreasing for ten years after disturbance before recovering
to pre-disturbance values. The maximum richness loss or gain after
disturbance was positively correlated with the disturbance intensity
(\(\rho_{Spearman}^{Richness}=0.50\)). In all disturbed plots the
Simpson diversity first increased until a maximum reached after around
20 years. This maximum was positively correlated with the disturbance
intensity (\(\rho_{Spearman}^{Simpson}=0.77\)). The Simpson diversity
then stabilized except for two T3 plots (plots 8 and 12) for which it
kept increasing (Fig. \ref{fig:DivTaxo}).

@@

The plot 7 from treatment 1 displayed constantly outlying functional
richness and Rao diversity and was removed from the graphical
representation for better readability. In undisturbed plots both
functional richness and Rao diversity remained stable along the 30
years. In disturbed plots, both trajectories depended on the disturbance
intensity, with their maximum being positively correlated to \%AGB loss
\(\rho_{Spearman}^{Richness}=0.76\) and \(\rho_{Spearman}^{Rao}=0.60\).
Functional richness and Rao diversity displayed for low disturbance
intensity a low but long-lasting increase up to a maximum reached after
20-25 years, and for high intensity, a fast but short increase followed
after 10 years by a slow decrease towards the initial values.

@@

Before disturbance, the different communities had different taxonomic
compositions, visualized by their distinct starting points on the NMDS
axis 2. These initial differences were maintained all along the 30 years
following disturbance, with post-disturbance trajectories corresponding
to a displacement on the NMDS axis 1. The taxonomic compositional
changes following disturbance were similar among communities, which may
correspond to the recruitment of a group of pioneers common to all
communities whenever their initial differences and the disturbance
intensity \citep{Denslow2000, Bongers2009}. Taxonomic trajectories
continued with a recovery of the initial composition which, althought
not fully achieved after 30 years, suggested the community taxonomic
resilience and the maintenance of community initial differences
\citep{Folke2006}. The return towards initial taxonomic composition
showed that species not belonging to the pre-disturbance community were
rarely recruited, probably because of the common dispersal limitations
among tropical tree species \citep{Svenning2005}.

Regarding functional composition, initial communities were similar and
trajectories were confounded in the functional space. Following
disturbance, the functional composition of pre-disturbance surviving
trees is believed to be the same as this of the initial community
\citep{Herault2018}. Post-disturbance changes in functional composition
then relied upon the recruitment of species or functional types
previously infrequent or absent. These species were probably competitive
pioneers becoming dominant as the most efficient to benefit from the
light, space and nutrients made available by disturbance. This
recruitment of pioneers drove functional composition trajectories
similar for all plots and disturbance intensity. The composition first
went towards more resource-acquisitive strategies, which was translated
by a displacement to the right along the first axis in the functional
plane \citep{Westoby1998, Wright2004, Reich2014}. Thereafter, the
pioneers recruited primarily were progressively excluded by long-lived,
more resistant and shade-tolerant species. The community functional
composition then returned towards more resource-conservative strategies,
suggesting the recovery of the initial community composition and
translated in the functional plane by a displacement left along the
first axis and upward along the second axis (Fig. \ref{fig:NMDSplans}).

Both taxonomic and functional composition trajectories initiated a
return towards pre-disturbance state after 30 years, which highlighted
the taxonomic and functional resilience of communities. There was
however a decoupling between taxonomic and functional composition
trajectories: while as taxonomic trajectories maintained the initial
differences among communities, the functional trajectories where similar
and convergent in the functional space \citep{Fukami2005}.

@@

Community taxonomic richness and evenness trajectories were determined
by the disturbance intensity, ranging from a limited and temporary
impact to significant and persistant alterations of community diversity.
The taxonomic trajectories markedly changed above an intensity threshold
for which the taxonomic richness was maximized and the taxonomic
evenness remained resilient. The disturbance intensity determined the
balance in the community between pre-disturbance surviving trees, and
trees recruited afterward. Below a 20\%-25\% AGB loss, the trees
surviving after disturbance remained numerous enough to maintain the
high taxonomic richness of the pre-disturbance community
\citep{Bongers2009}. The recruitment of pioneers, infrequent or absent
before disturbance, then increased the taxonomic richness all the more
so that the disturbance was intense \citep{Martin2015, Chaudhary2016}.
As these pioneers became more dominant they balanced the usual
hyper-dominance of tropical forests, hence temporarily increasing the
taxonomic evenness \citep{Baraloto2012a}. Above the intensity threshold,
the taxonomic richness of surviving trees was too low to be offset by
the recruitment of pioneers. In the Guiana shield indeed, the pool of
true pioneers specifically recruited after disturbance is restricted to
a few common genera (e.g. \emph{Cecropia} spp., \emph{Vismia} spp.)
\citep{Guitet2018}. The taxonomic richness then decreased following
intense disturbance, all the more so that the disturbance intensity was
high \citep{Molino2001}. At the same time, pioneers became persistently
dominant and prevented a return towards the initial evenness and the
recovery of hyper-dominant shade-tolerant species.

Taxonomic trajectories following disturbance of intermediate intensity
besides highlighted an intermediate time after disturbance, around 15-20
years, for which the taxonomic evenness was maximized. As already
observed in the Guiana shield \citep{Baraloto2012a} and in Bornean
tropical forests \citep{Cannon1998}, taxonomic evenness followed
hump-shaped trajectories with a peak when the recruitment was balanced
between pioneers and late-successional species. This intermediate time
corresponded to the match between the generation time of pionneers,
first to dominate the post-disturbance recruitment, and the time for
competitive exclusion to emerge, which leads to the dominance of
late-successional species.

Regarding community functional trajectories in contrast, there was
neither intermediate disturbance intensity nor intermediate time after
disturbance, which dismissed the IDH. Neither functional richness nor
functional evenness displayed a humped-shaped pattern, but both rather
increased with the recruitment of pioneers that were functionally highly
different from the pre-disturbance community
\citep{Denslow1980, Molino2001}. Above the intensity threshold however,
for the most intense disturbance, community functional richness and
evenness started to decrease after 15 to 20 years. The right after
disturbance, short-lived species benefited from the resources made
available and prevented the establishment of other species. The decline
of these short-lived pioneers decreased the functional richness and
evenness, but we suggest that the establishment of long-lasting pioneers
will follow and that the trajectories will catch up with those observed
for intermediate disturbance \citep{Walker2009}.

The IDH then translated into community taxonomic response, with
post-disturbance trajectories markedly different below and above an
intensity threshold. The IDH was also valid at an intermediate time
after disturbance, when the recruitment of early- and late-successional
species were balanced. Functional trajectories, however, remained
decoupled from taxonomic ones and dismissed the IDH.

@@

The decoupling between taxonomic and functional trajectories was
explained by a decrease in the functional redundancy within the
pre-disturbance functional space, due to the loss of species following
disturbance. The recruitment of pioneers, functionally different from
the pre-disturbance functional composition, did not compensate the
decrease of functional redundancy in the first place. Progressively
though, the functional redundancy was restored through the replacement
of the first established species by more competitive, long-lived
pioneers or late-successional species that were functionally closer to
the pre-disturbance community. This replacement followed the lottery
recruitment rules, implying an easy recruitment for the first species
but becoming increasingly hampered by the emergence of interspecific
competition \citep{Busing2002}. The recovery of the functional
redundancy then relied upon the random processe of species recruitment
and was increasingly slow and difficult to anticipate
\citep{Elmqvist2003, Diaz2005}.

@@

Community post-disturbance trajectories in compositon and diversity were
shaped by the recruitment of a determined pool of pioneers identical
among communities and disturbance intensity. Nevertheless, taxonomic and
functional composition trajectories were decoupled: while functional
trajectories remained similar in the functional space and converged
towards the recovery of a comparable initial state, taxonomic
trajectories shiwed initial differences in community composition that
were maintained along time. This decoupling was explained by a decrease
of the functional redundancy which mitigated the functional impact of
disturbance. Community functional and taxonomic diversity response to
disturbance was contrasted as well. While the functional trajectories
remained similar whenever the disturbance intensity, taxonomic
trajectories were markedly different above a threshold of 20-25\% AGB
removed for which taxonomic richness was maximized and taxonomic eveness
remained resilient. The Intermediate Disturbance Hypothesis applied to
the linkage of disturbance with the taxonomic diversity, but not with
the functional diversity. Taxonomic diversity was maximized following a
disturbance of intermediate intensity, that was the threshold of 25\%
AGB removed, and at an intermediate time after disturbance, that was
around 25 years after disturbance when the recruitment of early- and
late-successional species were balanced. Whenever the disturbance
intensity, community resilience (in terms of recovery of the
pre-disturbance state) was tangible but required several decades and
relied upon the random, lottery recruitment of rare species. Although
resilient, the functional redundancy was lower for more than 30 years
after disturbance which cautioned against the risks of infrequent
species loss and the persistence of disturbance-specific communities
\citep{Herault2018}.

>>>>

While both taxonomic and functional composition remained stable in
undisturbed communities (Fig. \ref{fig:NMDSplans}), they followed marked
and consistent trajectories over post-\break disturbance time. In
disturbed communities the taxonomic composition changed towards more
species with acquisitive functional strategies. This translated into a
switch from high mean WSG in the community to high mean SLA and
chlorophyll content (see appendix I). The functional composition of
disturbed communities corresponded to cyclic compositional changes with
an incomplete recovery of the initial composition (Fig.
\ref{fig:NMDSplans}).

The maximum dissimilarity with the initial state was positively
correlated with the disturbance intensity for both taxonomic and
functional composition (\(\rho_{Spearman}^{Taxonomic}=0.87\) and
\(\rho_{Spearman}^{Functional}=0.90\) respectively). The maximum
dissimilarity with the initial state was reached for taxonomic
composition between 15 to 25 years, for most of the plots, and around 22
years for functional composition.

@@

Community CWM average value of Maximum height at adult stage
(\emph{Hmax}), leaf toughness and wood specific gravity (\emph{WSG})
decreased and then slightly increased, but remained significantly lower
than their initial value (Fig. \ref{fig:CWM}). On the other side, bark
thickness and specific leaf area (\emph{SLA}) both increased. Bark
thickness remained substantially high after 30 years, and \emph{SLA} had
almost recovered to its initial value. For all traits, the maximum
difference to initial value was correlated to the disturbance intensity.
Positive correlations were observed for Leaf thickness, chlorophyll
content, SLA and bark thickness
(\(\rho_{Spearman}^{Leaf thickness}=0.76\),
\(\rho_{Spearman}^{Chlorophyll content}=0.60\),
\(\rho_{Spearman}^{SLA}=0.93\),
\(\rho_{Spearman}^{Bark thickness}=0.71\)). Negative correlation was
observed for Leaf toughness, WSG and Hmax
(\(\rho_{Spearman}^{Leaf toughness}=-0.53\),
\(\rho_{Spearman}^{WSG}=-0.75\), \(\rho_{Spearman}^{Hmax}=-0.40\)) The
proportions of the three lightest seed mass classes increased in all
disturbed plots. After 30 years the proportion of lightest seed mass
class decreased while it stabilized for the two other lightest seed mass
classes (Supp. Mat. - Fig. S2).

@@

For undisturbed plots, taxonomic richness and Simpson diversity remained
stable over the 30 years of monitoring. In disturbed communities, after
low disturbance intensity the taxonomic richness increased, reaching a
maximum gain of 14 botanical genera (plot 3 from treatment 2). After
intense disturbance the taxonomic richness followed a more complex
trajectory, decreasing for ten years after disturbance before recovering
to pre-disturbance values. The maximum richness loss or gain after
disturbance was positively correlated with the disturbance intensity
(\(\rho_{Spearman}^{Richness}=0.50\)).

In all disturbed plots the Simpson diversity first increased until a
maximum reached after around 20 years. This maximum was positively
correlated with the disturbance intensity
(\(\rho_{Spearman}^{Simpson}=0.77\)). The Simpson diversity then
stabilized except for two T3 plots (plots 8 and 12) for which it kept
increasing (Fig. \ref{fig:DivTaxo}).

@@

The plot 7 from treatment 1 displayed constantly outlying functional
richness and Rao diversity and was removed from the graphical
representation for better readability. In undisturbed plots both
functional richness and Rao diversity remained stable along the 30
years. In disturbed plots, both trajectories depended on the disturbance
intensity, with their maximum being positively correlated to \%AGB loss
\(\rho_{Spearman}^{Richness}=0.76\) and \(\rho_{Spearman}^{Rao}=0.60\).
Functional richness and Rao diversity displayed for low disturbance
intensity a low but long-lasting increase up to a maximum reached after
20-25 years. Following high intensity, they displayed a fast but short
increase followed after 10 years by a slow decrease towards the initial
values.

@@

Before disturbance, the different communities had different taxonomic
compositions, visualized by their distinct starting points on the NMDS
axis 2. These initial differences were maintained all along the 30 years
following disturbance, with the disturbance leading a displacement on
the NMDS axis 1. The taxonomic compositional changes following
disturbance were similar among communities, which may correspond to the
recruitment of a group of pioneers common to all communities whenever
their initial differences and the intensity of disturbance
\citep{Denslow2000, Bongers2009}. Taxonomic trajectories continued with
a recovery of the initial composition which, althought not fully
achieved after 30 years, suggested the resilience of the taxonomic
composition resilience and the maintenance of the initial differences
among communities \citep{Folke2006}. The return towards initial
taxonomic composition showed that species not belonging to the
pre-disturbance community were rarely recruited, probably because of the
common dispersal limitations among tropical tree species
\citep{Svenning2005}.

Regarding functional composition, initial communities were similar and
trajectories were confounded in the functional space. Following
disturbance, the functional composition of pre-disturbance surviving
trees is believed to be the same as this of the initial community
\citep{Herault2018}. The changes in functional composition following
disturbance then relied upon the recruitment of species or functional
types previously infrequent or absent. These species were probably
competitive pioneers becoming dominant as the most efficient to benefit
from the light, space and nutrients made available by the disturbance.
This recruitment of pioneers led similar functional composition
trajectories for all plots, whenever the disturbance intensity. The
composition first went towards more resource-acquisitive strategies,
which was translated by a displacement to the right along the first axis
in the functional plane \citep{Westoby1998, Wright2004, Reich2014}.
Thereafter, the pioneers recruited primarily were progressively excluded
by long-lived, more resistant and shade-tolerant species. The community
functional composition then returned towards more resource-conservative
strategies, suggesting the recovery of the initial functional
composition. This recovery translated in the functional plane by a
displacement left along the first axis and upward along the second axis
(Fig. \ref{fig:NMDSplans}).

Both taxonomic and functional composition trajectories initiated a
return towards pre-disturbance state after 30 years, which highlighted
the taxonomic and functional resilience of communities. Taxonomicand
functional trajectories however appeared decoupled: while taxonomic
trajectories maintained the initial differences among communities, the
functional trajectories where similar and convergent in the functional
space \citep{Fukami2005}.

@@

Community taxonomic richness and evenness trajectories were determined
by the disturbance intensity, ranging from a limited and temporary
impact to significant and persistant alterations of community diversity.
The taxonomic trajectories markedly changed above an intensity threshold
at which the taxonomic richness was maximized and the taxonomic evenness
remained resilient. The disturbance intensity impacted the balance in
the community between pre-disturbance surviving trees, and trees
recruited afterward. On the one hand, below a 20\%-25\% AGB loss, the
trees surviving after disturbance remained numerous enough to maintain
the high taxonomic richness of the pre-disturbance community
\citep{Bongers2009}. The recruitment of pioneers, infrequent or absent
before disturbance, then increased the taxonomic richness all the more
so that the disturbance was intense \citep{Martin2015, Chaudhary2016}.
As these pioneers became more dominant they balanced the usual
hyper-dominance of tropical forests, hence temporarily increasing the
taxonomic evenness \citep{Baraloto2012a}. On the other hand, above the
intensity threshold, the taxonomic richness of surviving trees was too
low to be offset by the recruitment of pioneers. In the Guiana shield
indeed, the pool of true pioneers specifically recruited after
disturbance is restricted to a few common genera (e.g. \emph{Cecropia}
spp., \emph{Vismia} spp.) \citep{Guitet2018}. The taxonomic richness
then decreased following intense disturbance, all the more so that the
disturbance intensity was high \citep{Molino2001}. At the same time,
pioneers became persistently dominant and prevented a return towards the
initial evenness and the recovery of hyper-dominant shade-tolerant
species.

Taxonomic trajectories following disturbance of intermediate intensity
besides highlighted an intermediate time after disturbance, around 15-20
years, for which the taxonomic evenness was maximized. As already
observed in the Guiana shield \citep{Baraloto2012a} and in Bornean
tropical forests \citep{Cannon1998}, taxonomic evenness followed
hump-shaped trajectories with a peak when the recruitment was balanced
between pioneers and late-successional species. This intermediate time
corresponded to the match between the generation time of pionneers,
dominating first the post-disturbance recruitment, and the time for
competitive exclusion to emerge, leading afterwards to the dominance of
late-successional species.

Regarding community functional trajectories in contrast, there was
neither intermediate disturbance intensity nor intermediate time after
disturbance, which dismissed the IDH. Neither functional richness nor
functional evenness displayed a humped-shaped pattern, but both rather
increased with the recruitment of pioneers that were functionally highly
different from the pre-disturbance community
\citep{Denslow1980, Molino2001}. Above the intensity threshold however,
for the most intense disturbance, community functional richness and
evenness started to decrease after 15 to 20 years. Right after
disturbance, short-lived species benefited from the resources made
available and prevented the establishment of other species. The decline
of these short-lived pioneers decreased the functional richness and
evenness, but we suggest that the establishment of long-lasting pioneers
will follow and that the trajectories will catch up with those observed
for intermediate disturbance \citep{Walker2009}.

The IDH then translated into community taxonomic response, with
post-disturbance trajectories changing markedly below and above an
intensity threshold of 25 \% AGB loss. The IDH was also valid at an
intermediate time after disturbance, when the recruitment of early- and
late-successional species were balanced. Functional trajectories,
however, remained decoupled from taxonomic ones and dismissed the IDH.

@@

The decoupling between taxonomic and functional trajectories was
explained by a decrease in the functional redundancy within the
pre-disturbance functional space, due to the loss of species following
disturbance. The recruitment of pioneers, functionally different from
the pre-disturbance functional composition, did not compensate the
decrease of functional redundancy in the first place. Progressively
though, the functional redundancy was restored through the replacement
of the first established species by more competitive, long-lived
pioneers or late-successional species that were functionally closer to
the pre-disturbance community. This replacement followed the lottery
recruitment rules, implying an easy recruitment for the first species
but becoming increasingly hampered by the emergence of interspecific
competition \citep{Busing2002}. The recovery of the functional
redundancy then relied upon the random process of species recruitment
and was increasingly slow and difficult to anticipate
\citep{Elmqvist2003, Diaz2005}.

@@

Community post-disturbance trajectories in compositon and diversity were
shaped by the recruitment of a determined pool of pioneers identical
among communities and disturbance intensity. Nevertheless, taxonomic and
functional composition trajectories appeared decoupled. While functional
trajectories remained similar in the functional space and converged
towards the recovery of a comparable initial state, taxonomic
trajectories showed initial differences in community composition that
were maintained along time. This decoupling was explained by a decrease
of the functional redundancy that mitigated the functional impact of
disturbance. Community functional and taxonomic diversity response to
disturbance was contrasted as well. While the functional trajectories
remained similar whenever the disturbance intensity, taxonomic
trajectories were markedly different above a threshold of 20-25\% AGB
removed for which taxonomic richness was maximized and taxonomic eveness
remained resilient. The Intermediate Disturbance Hypothesis applied to
the linkage of disturbance with the taxonomic diversity, but not with
the functional diversity. Taxonomic diversity was maximized following a
disturbance of intermediate intensity, that was the threshold of 25\%
AGB removed, and at an intermediate time after disturbance, that was
around 25 years after disturbance when the recruitment of early- and
late-successional species were balanced. Whenever the disturbance
intensity, community resilience (in terms of recovery of the
pre-disturbance state) was tangible but required several decades and
relied upon the random lottery recruitment of rare species. Although
resilient, the functional redundancy was lower for more than 30 years
after disturbance which cautioned against the risks of infrequent
species loss and the persistence of disturbance-specific communities
\citep{Herault2018}.
